---
title: "Manu-stats"
author: "Gina Nichols"
date: "December 27, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = "../")
library(ezknitr)
library(lme4)
library(tidyverse)
library(lmerTest)
library(janitor)
library(emmeans)

dat <- read_csv("../_tidydata/td_cc-database-clean-long.csv")

den <- dat %>% filter(resp == "den") %>% filter(!is.na(LRR)) %>% mutate(cc_bio_Mgha = cc_bio_kgha/1000)
bio <- dat %>% filter(resp == "bio") %>% filter(!is.na(LRR)) %>% mutate(cc_bio_Mgha = cc_bio_kgha/1000)

```

# Overall, what is the mean cover crop biomass for each cover crop group (grass vs non-grass)?
```{r mean ccbiomass}
m1 <- (lmer(cc_bio_Mgha ~ cc_type2 + (1|study), data = bio))
summary(m1)
```

# In which modifiers do we need to include CC biomass as a covariate? 
```{r ccbiomass cov}

# msmt_season, no
summary(lmer(cc_bio_Mgha ~ msmt_season + (1|study), data = bio))

# weed_group, no
summary(lmer(cc_bio_Mgha ~ weed_group + (1|study), data = bio))

# sys_tillage, no
summary(lmer(cc_bio_Mgha ~ cropsys_tillage + (1|study), data = bio))

# aridity, no
summary(lmer(cc_bio_Mgha ~ aridity_index + (1|study), data = bio))

# OM, yes, but it goes down as OM pct goes up. I guess include it?
summary(lmer(cc_bio_Mgha ~ OM_pct + (1|study), data = bio))

```

# What is the overall effect of CCs on weed biomass and density?
Do CCs sig effect weed *biomass*?
```{r overall bio}
  
mo.bio <- lmer(LRR ~ 1 + (1|study), data = bio, weights = wgt)
summary(mo.bio)

```
Yes, what are the estimates and 95% confidence ranges?
```{r overall bio est}

as_tibble(contest(mo.bio, L=1, joint=F, level=0.95)) %>% 
  clean_names() %>% 
  select(estimate, lower, upper, pr_t) %>% 
  gather(key = parm, value = val) %>% 
  mutate(val_pct = 100 - exp(val)*100)

```
What is the fail-safe-number for this significant result?
```{r fsn bio}
library(metafor)
bio_fsn <- bio %>%
  mutate(yi = LRR,
         vi = 1/wgt)

fsn(yi, vi, data=bio_fsn)
```
Do CCs sig effect weed *density*?
```{r overall den}
  
mo.den <- lmer(LRR ~ 1 + (1|study), data = den, weights = wgt)
summary(mo.den)
```

# Which modifiers have significantly different values?
```{r sig mods bio}
read_csv("../_tidydata/sd_mods-weeds-contrasts.csv") %>% 
  filter(resp == "bio") %>% 
  mutate_if(is.numeric, round, 3) %>% 
  select(resp, mod_code, level1, level2, est, se, df, p.value) %>% 
  arrange(resp, p.value)
```
1. Measurement season
1. Measurement in reference to planting
1. CC type
1. weed group

```{r sig mods den}
read_csv("../_tidydata/sd_mods-weeds-contrasts.csv") %>% 
  filter(resp == "den") %>% 
  mutate_if(is.numeric, round, 3) %>% 
  select(resp, mod_code, level1, level2, est, se, df, p.value) %>% 
  arrange(resp, p.value)
```
1. Measurement season (but only bc it includes fall)
1. Following crop (only because we have corn/soy)
1. and MAYBE cropsys_tillage

For the significant modifiers, what are the estimates and CIs for weed *biomass*?
```{r sig mods bio est}
read_csv("../_tidydata/sd_mods-weeds-contrasts.csv") %>% 
  filter(resp == "bio") %>% 
  filter(p.value < 0.05) %>% 
  mutate_if(is.numeric, round, 3)

read_csv("../_tidydata/sd_mods-weeds.csv") %>% 
  select(resp, modlvl, est) %>%  
  mutate(val_pct = 100-exp(est)*100)
  
```