---
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(lme4)
```

# Standard Error for Inverse Prediction with Random Effect Models 

AES Statistical Consulting  
Consultant: Katherine Goode  
Client: Gina Nichols, graduate student in agronomy (vnichols@iastate.edu)  
PI:  Matt Liebman (mliebman@iastate.edu)  
`r format(Sys.time(), '%B %d, %Y')`  

## Data, Model, and Inverse Prediction

Gina is working on a meta analysis where the response variable is the log response ratio (LRR) associated with yield. She is interested in comparing this to weed biomass. As is typical in a meta-analysis, she has data from various studies. I do not have access to the data, so I generated an example dataset to work with in this document. A plot of this data is shown below.

\vspace{0.25cm}

```{r echo = FALSE, fig.align = "center", fig.width = 6.5, fig.height = 4}
# Generate the example data
set.seed(20191118)
ex_data <- data.frame(lrr = c(runif(20, -5, 5),
                              runif(20, -5, 5),
                              runif(20, -5, 5),
                              runif(20, -5, 5),
                              runif(20, -5, 5)),
                      study = rep(c("Study 1", 
                                    "Study 2", 
                                    "Study 3",
                                    "Study 4",
                                    "Study 5"),
                                  each = 20)) %>%
  arrange(study, lrr) %>%
  mutate(lrr = lrr + rnorm(100, 0, 2),
         weed_biomass = c(sort(runif(20, 30, 50)),
                          sort(runif(20, 30, 50)),
                          sort(runif(20, 30, 50)),
                          sort(runif(20, 30, 50)),
                          sort(runif(20, 30, 50)))) %>%
  mutate(study = factor(study))

# Plot the example data
ex_data %>%
  ggplot(aes(x = weed_biomass, lrr)) + 
  geom_point() + 
  facet_wrap(. ~ study) + 
  labs(x = "Weed Biomass", y = "Log Response Ratio")
```

\vspace{0.25cm}

The model that Gina is using has the form

$$\mbox{LRR}_i=\beta_0+\beta_1\mbox{weed biomass}_i+\alpha_i+\epsilon_i$$

where $\alpha_i\sim N\left(0,\sigma_{study}^2\right)$ is a study random effect and $\epsilon_i\sim N\left(0,\sigma_{error}^2\right)$ is the error term for $i=1,...,n$. I fit this model to the example data. The code and summary of the model are included below.

\vspace{0.25cm}

```{r}
m <- lmer(lrr ~ weed_biomass + (1|study), data = ex_data)
summary(m)
```

\vspace{0.25cm}

Gina is interested in predicting the weed biomass given a LRR. In particular, she is interested in the case when LRR = -0.69, which corresponds to 50% weed control. Additionally, she would like the standard error associated with the predicted weed biomass. The estimate of the weed biomass for a given LRR can be computed using inverse prediction as

$$\hat{p}_{\mbox{weed control}} = \frac{\mbox{LRR} - \hat{\beta}_0}{\hat{\beta}_1}.$$
In order to obtain the standard error for $\hat{p}_{\mbox{weed control}}$, we will use the delta method.

## Delta Method

### Overview of the Delta Method Theory

Adapted from Dr. Mark Kaiser's Stat 520 course notes from fall 2016: 

\begin{itemize}
\item Let $\hat{\boldsymbol{\theta}}_n$ be a sequence of asymptotically normal estimators of a parameter vector $\boldsymbol{\theta}=(\theta_1,...,\theta_p)$ with mean $\boldsymbol{\theta}$ and covariance matrix $c_n^2\boldsymbol{\Sigma}$.

\item Let $g(\boldsymbol{\theta})$ be a real-valued function of $\boldsymbol{\theta}$ that is continuously differentiable in a neighborhood of $\boldsymbol{\theta}$.

\item Let $\textbf{d}$ be a vector of length $p$ with a $j$th element of

$$\frac{\partial g(\boldsymbol{\theta})}{\partial \theta_j}.$$

\item Then $g\left(\hat{\boldsymbol{\theta}}\right)$ is asymptotically normal with mean $g(\boldsymbol{\theta})$ and covariance matrix $c_n^2\textbf{d}\boldsymbol{\Sigma}\textbf{d}'$.

\item In likelihood estimation and inference, $\hat{\boldsymbol{\theta}}_n$ can be used as a plug-in estimator of $\boldsymbol{\theta}$ to estimate $g(\boldsymbol{\theta})$ and $c_n^2\textbf{d}\boldsymbol{\Sigma}\textbf{d}'$.

\item Additionally, in the iid case, $c_n^2\boldsymbol{\Sigma}$ is the inverse total information matrix. That is 

$$c_n^2\boldsymbol{\Sigma}=\frac{1}{n}I(\boldsymbol{\theta})^{-1}.$$

This can be estimated by the negative inverse Hessian matrix, which is the estimated variance-covariance matrix of $\boldsymbol{\theta}$.

\item Thus, the variance of $g\left(\hat{\boldsymbol{\theta}}\right)$ can be obtained as

$$Var\left(g\left(\hat{\boldsymbol{\theta}}\right)\right)=\textbf{d}Cov\left(\hat{
\boldsymbol{\theta}}\right)\textbf{d}',$$

and the standard error of $g\left(\hat{\boldsymbol{\theta}}\right)$ can be computed as

$$SE\left(g\left(\hat{\boldsymbol{\theta}}\right)\right)=\sqrt{Var\left(g\left(\hat{\boldsymbol{\theta}}\right)\right)}.$$

\end{itemize}

A good reference on the delta method that also describes how to apply the delta method in R can be found on the [IDRE website](https://stats.idre.ucla.edu/r/faq/how-can-i-estimate-the-standard-error-of-transformed-regression-parameters-in-r-using-the-delta-method/).

### Notation and Derivatives for the Weed Biomass Problem

In our case, $\boldsymbol{\theta}=\left(\beta_0, \beta_1\right)$ and

$$g(\boldsymbol{\theta}) = \frac{\mbox{LRR} - \beta_0}{\beta_1}.$$

To obtain $\textbf{d}$, we need to compute the partial derivatives of $g(\boldsymbol{\theta})$ in terms of $\beta_0$ and $\beta_1$. These can be computed as follows.

\begin{align*}
\frac{\partial g(\boldsymbol{\theta})}{\partial\beta_0} & = \frac{\partial}{\partial \beta_0}\left(\frac{LRR - \beta_0}{\beta_1}\right)\\
& = \frac{\partial}{\partial \beta_0}\left(\frac{LRR}{\beta_1}\right) - \frac{\partial}{\partial \beta_0} \left(\frac{\beta_0}{\beta_1}\right)\\
& = 0 - \frac{1}{\beta_1}\\
& = -\frac{1}{\beta_1}
\end{align*}

\begin{align*}
\frac{\partial g(\boldsymbol{\theta})}{\partial \beta_1} & = \frac{\partial}{\partial \beta_1} \left(\frac{LRR - \beta_0}{\beta_1}\right)\\
& = \frac{\partial}{\partial \beta_1} \left(LRR - \beta_0\right)(\beta_1^{-1})\\
& = -\left(LRR - \beta_0\right)(\beta_1^{-2})\\
& = \frac{-LRR+\beta_0}{\beta_1^2}
\end{align*}

Thus,

\begin{align}
\textbf{d} = \left[\begin{array}{cc} -\frac{1}{\beta_1} & \frac{-LRR+\beta_0}{\beta_1^2}\end{array}\right]
\end{align}

The standard error for $g(\hat{\boldsymbol{\theta}})$ will be

$$\sqrt{Var[g(\hat{\boldsymbol{\theta}})]}=\left(\textbf{d}Cov\left[\hat{\boldsymbol{\theta}}\right]\textbf{d}'\right)^{1/2}$$
where

$$Cov\left[\hat{\boldsymbol{\theta}}\right]=\left[\begin{array}{cc} 
Var\left[\hat{\beta}_0\right] & Cov\left[\hat{\beta}_0, \hat{\beta}_1\right] \\
Cov\left[\hat{\beta}_0, \hat{\beta}_1\right] & Var\left[\hat{\beta}_1\right] 
\end{array}\right].$$
  
## Applying the Delta Method to the Inverse Prediction

### R Function 

I wrote the function `compute_se` to implement the delta method computations to compute the standard error for the weed biomass prediction for a given LRR. The function also returns the estimate of the LRR and a 95% confidence interval for the prediction. The inputs and outputs of the function are as follows.

Inputs:

* `lrr`: LRR for which to compute the weed biomass
* `betas`: estimated regression coefficients of $\beta_0$ and $\beta_1$ (should be a vector of length 2)
* `vcov`: estimated variance covariance matrix of $\beta_0$ and $\beta_1$ (should be a 2x2 matrix)

Outputs:

* data frame with the variables of 
    + `lrr`: age that was specified for the computations
    + `pred_biomass`: estimated weed biomass for the specified `lrr`
    + `se`: standard error for the estimated weed biomass (computed using the delta method)
    + `ci_Lower`: lower bound of the 95% confidence interval for weed biomass
    + `ci_Upper`: upper bound of the 95% confidence interval for weed biomass

The code for the function `compute_se` is included below.

\vspace{0.25cm}

```{r}
# Function for computing the delta-method standard error of weed biomass
compute_se <- function(lrr, betas, vcov){
  
  # Separate the betas
  b0 <- betas[1] 
  b1 <- betas[2] 
  
  # Compute the inverse prediction of weed biomass
  pred_biomass <- (lrr - b0) / b1

  # Create an empty 1x2 matrix to store the elements of d in
  d <- matrix(NA, nrow = 1, ncol = 2)
  
  # Compute the elements of d (partial derivatives of g(beta))
  d[1] <- -1 / b1 
  d[2] <- (-lrr + b0) / (b1^2) 
  
  # Compute the standard error of annual survival (using the delta method)
  se <- sqrt(d %*% vcov %*% t(d))

  # Compute the lower and upper bounds of the 95% CI for annual survival
  lower <- pred_biomass - (1.96 * se) 
  upper <- pred_biomass + (1.96 * se)

  # Return the log response ratio, the predicted weed biomass given the 
  # resopnse ration, the delta method standard error, and the lower 
  # and upper bounds of the 95% CI for the predicted weed biomass
  return(data.frame(lrr = lrr,
                    pred_biomass = pred_biomass, 
                    se = se, 
                    ci_lower = lower, 
                    ci_upper = upper))
  
}
```

\vspace{0.25cm}

### Standard Error Calculation for LRR = -0.69

I first extracted the vector of estimated betas and the variance-covariance matrix.

\vspace{0.25cm}

```{r}
# Extract the variance-covariance matrix for beta0 and beta1
betas <- as.vector(summary(m)$coefficients[,1])
betas

# Extract the variance-covariance matrix for beta0 and beta1
vcov <- matrix(vcov(m), nrow = 2)
vcov
```

\vspace{0.25cm}

Then I applied the `compute_se` function to compute the weed biomass estimate, standard error, and 95% confidence interval for a LRR of -0.69. The results are shown below.

\vspace{0.25cm}

```{r}
# Apply the compute_se function
res <- compute_se(lrr = -0.69, betas = betas, vcov = vcov)
res
```


