---
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(lme4)
```

# Standard Error for Inverse Prediction with Random Effect Models (no intercept model)

AES Statistical Consulting  
Consultant: Katherine Goode  
Client: Gina Nichols, graduate student in agronomy (vnichols@iastate.edu)  
PI:  Matt Liebman (mliebman@iastate.edu)  
`r format(Sys.time(), '%B %d, %Y')`  

## Data, Model, and Inverse Prediction

Gina is now fitting a model with no intercept. Again, I will work with the data I generated previously. The plot of the data is shown below.

\vspace{0.25cm}

```{r echo = FALSE, fig.align = "center", fig.width = 6.5, fig.height = 4}
# Generate the example data
set.seed(20191118)
ex_data <- data.frame(lrr = c(runif(20, -5, 5),
                              runif(20, -5, 5),
                              runif(20, -5, 5),
                              runif(20, -5, 5),
                              runif(20, -5, 5)),
                      study = rep(c("Study 1", 
                                    "Study 2", 
                                    "Study 3",
                                    "Study 4",
                                    "Study 5"),
                                  each = 20)) %>%
  arrange(study, lrr) %>%
  mutate(lrr = lrr + rnorm(100, 0, 2),
         weed_biomass = c(sort(runif(20, 30, 50)),
                          sort(runif(20, 30, 50)),
                          sort(runif(20, 30, 50)),
                          sort(runif(20, 30, 50)),
                          sort(runif(20, 30, 50)))) %>%
  mutate(study = factor(study))

# Plot the example data
ex_data %>%
  ggplot(aes(x = weed_biomass, lrr)) + 
  geom_point() + 
  facet_wrap(. ~ study) + 
  labs(x = "Weed Biomass", y = "Log Response Ratio")
```

\vspace{0.25cm}

The model that Gina is using has the form

$$\mbox{LRR}_i=\beta_1\mbox{weed biomass}_i+\alpha_i+\epsilon_i$$

where $\alpha_i\sim N\left(0,\sigma_{study}^2\right)$ is a study random effect and $\epsilon_i\sim N\left(0,\sigma_{error}^2\right)$ is the error term for $i=1,...,n$. I fit this model to the example data. The code and summary of the model are included below.

\vspace{0.25cm}

```{r}
m <- lmer(lrr ~ 0 + weed_biomass + (1|study), data = ex_data)
summary(m)
```

\vspace{0.25cm}

Gina is interested in predicting the weed biomass given a LRR. In particular, she is interested in the case when LRR = -0.69, which corresponds to 50% weed control. Additionally, she would like the standard error associated with the predicted weed biomass. The estimate of the weed biomass for a given LRR can be computed using inverse prediction as

$$\hat{p}_{\mbox{weed control}} = \frac{\mbox{LRR}}{\hat{\beta}_1}.$$
In order to obtain the standard error for $\hat{p}_{\mbox{weed control}}$, we will use the delta method.

# Notation and Derivatives for the Weed Biomass Problem

In this case, $\boldsymbol{\theta}=\left(\beta_1\right)$ and

$$g(\boldsymbol{\theta}) = \frac{\mbox{LRR}}{\beta_1}.$$

To obtain $\textbf{d}$, we need to compute the derivative of $g(\boldsymbol{\theta})$ in terms of $\beta_1$. This can be computed as follows.

$$\begin{aligned}
\frac{d g(\boldsymbol{\theta})}{d \beta_1} & = \frac{d}{d \beta_1} \left(\frac{LRR}{\beta_1}\right)\\
& = \frac{d}{d \beta_1} \left(LRR\right)(\beta_1^{-1})\\
& = -\left(LRR\right)(\beta_1^{-2})\\
& = \frac{-LRR}{\beta_1^2}
\end{aligned}$$

Thus,

$$\begin{aligned}
\textbf{d} = \left[\begin{array}{cc}\frac{-LRR}{\beta_1^2}\end{array}\right]
\end{aligned}$$

The standard error for $g(\hat{\boldsymbol{\theta}})$ will be

$$\sqrt{Var[g(\hat{\boldsymbol{\theta}})]}=\left(\textbf{d}Cov\left[\hat{\boldsymbol{\theta}}\right]\textbf{d}'\right)^{1/2}$$
where

$$Cov\left[\hat{\boldsymbol{\theta}}\right]=Var\left[\hat{\beta}_1\right].$$
  
## Applying the Delta Method to the Inverse Prediction

### R Function 

I wrote a new function `compute_se_noint` to implement the delta method computations to compute the standard error for the weed biomass prediction for a given LRR when there is no intercept in the model. The function also returns the estimate of the LRR and a 95% confidence interval for the prediction. The inputs and outputs of the function are as follows.

Inputs:

* `lrr`: LRR for which to compute the weed biomass
* `betas`: estimated regression coefficients of $\beta_1$ (should be a vector of length 1)
* `vcov`: estimated variance of $\beta_1$ (should be a vector of length 1)

Outputs:

* data frame with the variables of 
    + `lrr`: age that was specified for the computations
    + `pred_biomass`: estimated weed biomass for the specified `lrr`
    + `se`: standard error for the estimated weed biomass (computed using the delta method)
    + `ci_Lower`: lower bound of the 95% confidence interval for weed biomass
    + `ci_Upper`: upper bound of the 95% confidence interval for weed biomass

The code for the function `compute_se` is included below.

\vspace{0.25cm}

```{r}
# Function for computing the delta-method standard error of weed biomass
compute_se_noint <- function(lrr, betas, vcov){
  
  # Shorten the name of betas
  b1 <- betas 
  
  # Compute the inverse prediction of weed biomass
  pred_biomass <- lrr / b1

  # Compute d (partial derivatives of g(beta))
  d <- (-lrr) / (b1^2) 
  
  # Compute the standard error (using the delta method)
  se <- sqrt(d %*% vcov %*% t(d))

  # Compute the lower and upper bounds of the 95% CI
  lower <- pred_biomass - (1.96 * se) 
  upper <- pred_biomass + (1.96 * se)

  # Return the log response ratio, the predicted weed biomass given the 
  # resopnse ration, the delta method standard error, and the lower 
  # and upper bounds of the 95% CI for the predicted weed biomass
  return(data.frame(lrr = lrr,
                    pred_biomass = pred_biomass, 
                    se = se, 
                    ci_lower = lower, 
                    ci_upper = upper))
  
}
```

\vspace{0.25cm}

### Standard Error Calculation for LRR = -0.69

I first extracted the estimated and variance of $\beta_1$.

\vspace{0.25cm}

```{r}
# Extract the estimate of beta1
betas <- as.vector(summary(m)$coefficients[,1])
betas

# Extract the variance for beta1
vcov <- as.vector(vcov(m))
vcov
```

\vspace{0.25cm}

Then I applied the `compute_se_noint` function to compute the weed biomass estimate, standard error, and 95% confidence interval for a LRR of -0.69. The results are shown below.

\vspace{0.25cm}

```{r}
# Apply the compute_se function
res <- compute_se_noint(lrr = -0.69, betas = betas, vcov = vcov)
res
```


